---
title: "Investigation of the alive size of the HIV/AIDS epidemic from 1990 - 2020"
author: "Khushi Chauhan, khushic4@uw.edu; Wenqing Cao, wenqic2@uw.edu; Andy Cerekja, acerekja@uw.edu; Parker Huang, phuang20@uw.edu"
date: "2/18/2022"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
source("final-project-starter-wenqic2-1976285/source/Graph-1.R")
source("final-project-starter-wenqic2-1976285/source/Graph-2.R")
source("final-project-starter-wenqic2-1976285/source/table-p02.R")
source("final-project-starter-wenqic2-1976285/data/summary_script.R")
```

#### Affiliation 
AC-6 - INFO-201: Technical Foundations of Informatics - The Information School - University of Washington 

#### Keywords
Virology, disease transmission, disease prevalence, disease distribution

## Abstract
What we care about is the survival rate of patients infected with AIDS. This question is important because it can help scientists investigate the change pattern of this disease. AIDS usually have long incubation periods and serious complications. Investigation of the survival rate and time period can help us better understand the incubation periods. To address the question, we will focus on the number of alived patient and area distribution factors through 1990-2020 to analyze the disease.

## Introduction
For our project, we will be investigating how the size of the HIV/AIDS epidemic has changed from year to year between 1990 to 2020. We will do this by analyzing HIV/AIDS datasets from the World Health Organization, including one that shows the number of people living with HIV each year. We will also be considering how the mortality and infection rates of HIV/AIDS have changed over that same time period. Investigating the mortality rate and the time period will help us to better understand the incubation rates of HIV/AIDS.

## Design Situation
Project Framing: This project will analyze data frames from the World Health Organization (WHO) in order to better understand how the transmissibility of HIV/AIDS has changed over three decades. The data set we will be focusing on is “number of people living with HIV/AIDS,” with the WHO giving estimates by region. Human Values: This research is valuable due to the lives that are impacted by the analysis of this data. By evaluating how the proportion of people living with HIV/AIDS has changed over three decades, scientists can better understand how effective their treatments have been and how to improve the conditions of these individuals’ lifestyles. It may also provide clues to other factors affecting the presence and transmission of HIV/AIDS such as socioeconomic status and regional wealth. The analysis of specific regions with high fluctuation of the number of people living with HIV/AIDS could provide scientists and researchers with information that could be useful in geographic advantages of certain regions with respect to transmissibility. Direct/Indirect Stakeholders: The direct stakeholders in this project are those infected with the illness. Direct stakeholders of this project are also officials in charge of managing the HIV/AIDS treatment funding for particular hospitals and regions. Indirect Stakeholders in this project are family members of those infected with HIV/AIDS, the governments associated with certain regions of higher rates of infection. Harms & Benefits: Harms of such a research project are that they can help promote negative stereotypes of certain regions. Typically regions corresponding to a high infection rate would be expected to be negatively stereotyped. If the research is improperly conducted, it can lead to misleading correlations between region and proportion of infected individuals which could result in improper government regulations. Benefits of such a research project are that they can in fact provide insights into positive treatments of infected individuals and better government regulations for these individuals. For example, areas of lingering or increasing high infection rates could be provided with more federal or state level funding for treatments.

## Research Questions
A research question that we will be addressing is, how has the number of people living with HIV changed from 1990 to 2020? Another one of our research questions is how has the mortality and transmission rates of HIV/AIDS changed over the years? Our last research question is, how does age affect the transmission rate of HIV/AIDS? The mortality and transmission rates are important because they can help scientists investigate the change patterns of HIV/AIDS. All three of these questions are important because they will allow us to gain a better understanding of not only how the size of the HIV/AIDS epidemic changed over time, but also what caused the changes in size.

## The Dataset
There are four data files included in the data set and we choose to use [Number of people (all ages) living with HIV](https://apps.who.int/gho/data/node.main.620?lang=en). This data file reports the number of people who are infected by the hiv/aids and still alive. This data file has two versions, one is sorted by countries, the other is by WHO region. There are 30 columns and 7/171 rows (based on different row name). Each column is featured with the number of years from 1990 to 2020. The row name can be the name of each country or the name of the region divided by WHO (World Health Organization). Each cell includes the cumulative number of alive people infected by hiv/aids at the end of each year as well as the error range of the number provided. The disaggregation includes females of all ages. The proportion of living AIDS patients in the total number of AIDS infected population and male patients are not included in this dataset. The data was estimate by each countries and collected by UNAIDS (The Joint United Nations Programme on HIV and AIDS) and WHO, and will be published every year. Standard methods and tools were used to predict and estimate the pattern of HIV. Statistics on the number of AIDS infections and deaths are helpful for medical workers to study the harm of it, control the spread of AIDS within the scope and ensure social stability. WHO is funded by both assessed contributions and voluntary contributions. WHO member states directly benefit from this data, which they can analyze and calculate to make adjustments to AIDS prevention and control. Because the data was collected by the WHO and UNAIDS. As the most official health organization, they are responsible for the source and authenticity of the data. Moreover, these two organizations have professional data collection methods and prediction models, which can ensure the accuracy of data. In addition, each unit in the data file contains the error range of the data, to further ensure the authenticity of the data. After our group comes up with this topic, we first search the related information by using keywords such as “aids” “morality” and “database” etc. on Google. Then we filtered irrelevant and incredible information and chose the data set provided by WHO, which is a professional and official source of the data we want.

## Implications
Through our research, there will be numerous possible implications for health officials around the world after being given the opportunity to analyze our results and answers to our research questions. In general, we of course hope for this research to lead to better understanding of the HIV/AIDS epidemic. More specifically, we expect to be able to analyze our current treatment for HIV/AIDS and see how effective our advances in treatment have been in trying to slow the spread of it. After seeing the variance in the mortality rate of HIV/AIDS and the total number of global cases, this data can then be used to see in which years these numbers were down, allowing for health officials to then address further research into those years to see what about these years made the numbers decrease so much and how can that information be used to further slow down HIV/AIDS in the future.

## Limitations
While we are given access to an absolutely massive database of information on our topic, the database possesses solely numeric data. While being given the numbers about the HIV/AIDS epidemic is still very valuable information, the numbers being given without any global context as to what was occurring at the time does bring down the value of the data slightly. For example, knowing that the 2020 numbers for HIV/AIDS transmission were down from the year prior may look as though we are making good progress in our fight against the epidemic, but given global context it can be understood that a massive factor in that drop was because of the COVID-19 pandemic. To address this limitation, doing some further research into big global events in anomalous years will be beneficial in our research and further research done in the future.


## Summary


## Table
This table shows the total number of people living with HIV worldwide from 1990 to 2020. The original data included the cumulative total number of lives after infected by HIV for each year between 1990 and 2020 for each continent. We focused on the global total so that we could judge the overall trend of the disease. In addition, we added the amount of change in the table to help us more intuitively observe the growth pattern from year to year. The table is sorted by year, with each row representing a particular year. Columns contain three main features: original data, estimated data, and year-to-year changes. As can be seen from the table, although the number of new cases actually decreased significantly from the 20th century to the 21st century, the overall number of AIDS patients continued to increase. And by looking at the changes, it can be found that from 2000 to 2020, the annual increase is repeated, which means that AIDS has not been effectively controlled.

```{r global_lives}
kable(global_lives)
```

## Chart 1
This graph show how the global number of people with HIV increase from 1990 to 2020. This graph shows that the number of HIV cases increased pretty quickly from 1990 to about two thousand and then it slowed down significantly. This graph also show that around 2008 or 2009 the number cases started to pick up again, but did not get anywhere close to where it was in the 90s.


## Chart 2
This graph tracks the number of people living with HIV in specifically the egion of the Americas from the years 1990-2020. By looking at this bar chart, a steady upwards trend can clearly be seen. This shows that in this region, not enough advancements have been made in terms of protection from and medicine for HIV. Collecting data like this for a region rather than globally assists in providing more specific data to look at for each region to see what they can change.
